HO unification from object language to meta language

### 2024-03-29 paper/main.tex
tex update

$\eta$-reduction is an equivalence relation over terms where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expanded
version of $t$. The implementation of the \elpiIn{comp} relation given in
\cref{sec:compilation} compiles the \Fo{} terms $t_1 =$ \elpiIn{flam x\ fapp
[fuva A, x]} and $t_2 =$ \elpiIn{fcon "f"} into the \Ho terms $t'_1 =
$ \elpiIn{lam x\ uva A' x} and $t'_2 =$ \elpiIn{fcon "f"} with mapping
\mapping{\elpiIn{A}}{\elpiIn{A'}}{1}. However, if the oracle sets \elpiIn{A'} to the constant
\elpiIn{"f"}, the unification of $t'_1$ and $t'_2$ in the meta language will
fail even though $t_1 \Eo t_2$. The reason of this failure is attributed to the
fact that $t'_1 =$~\elpiIn{lam x\ app[con "f", x]} cannot be unified with $t'_2
=$~\elpiIn{con "f"} since the two terms have different rigid heads. We solve this
unification problem by adapting the \elpiIn{comp} relation such that it
recognizes \maybeeta subterms and replaces them with fresh \Ho variables. 
This link between the variable $v$ and the subterm $t$ is stored in what we
call \linketa which is an object with the following type

@@

\begin{invariant}[\linketa \rhs]
  The \rhs of a \linketa having the shape $\lambda x.F_x$
  where $F_x$ is a term not starting with the \elpiIn{lam} constructor.
  \label{inv:link-eta-right}
\end{invariant}

In the next three subsections we explain how we detect \maybeeta terms, how we
compile them and how \linketa are activated during the execution of the program
and provide justification for why \cref{inv:linklhs,inv:link-eta-right} remain true.

@@

Compiling term with \maybeeta terms forces us to determine if, $\lambda x. T_x$,
for any term $T$ having $x$ in scope, can be a $\eta$-expansion, i.e. under a
given substitution $\sigma$, we have $\sigma (\lambda x.T_x) = t$. This
\maybeeta detection is not a trivial operation as it may seems.
\\
\noindent
\begin{minipage}{.5\linewidth}
  \begin{gather}
    \lambda x. f \appsep A_x\\
    \lambda x. f \appsep x \appsep A_x
  \end{gather}
\end{minipage}% This must go next to `\end{minipage}`
\begin{minipage}{.5\linewidth}
  \begin{gather}
    \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}\\
    \lambda x. f \appsep A_x \appsep x \label{eq:lamdisc}
  \end{gather}
\end{minipage}

>>>>

$\eta$-reduction is an equivalence relation where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expansion of $t$. 
\printAll
  {{\elpiIn{flam x\ fapp [fuva A, x]} \Uo \elpiIn{fcon "f"}}}
  {{\lambda x. A'_x \Ue f}}
  {{\mapping{\elpiIn{A}}{A'}{1}}}
  {{}}
Following the implementation of \elpiIn{comp} given in
\cref{sec:compilation}, the unification problem $\mathcal{P}_1$ given 
above, is compiled into the unification problem $\mathcal{T}_1$. 
We can note that all the terms appearing in $\mathcal{P}_1$ are
unifiable by \Uo, but, despite this, \Ue, is not able to solve $\mathcal{T}_1$:
its left and right hand sides have both different rigid head:
$\lambda x.A'_x$ corresponds to the \Ho term \elpiIn{lam x\ app[con "f", x]} 
and $f$ corresponds to \elpiIn{con"f"}. 
This failure is motivated by the fact that \elpiIn{flam x\ fapp [fuva A, x]}
is a term belonging to \maybeeta.

In order to guarantee \cref{prop:simulation}, we need to modify the way terms are
compiled. The goal is to recognize
every \maybeeta subterm $t$ and replace it with fresh \Ho variables $v$. 
This connection between the variable $v$ and the subterm $t$ is stored in what we
call \linketa which is an object with the following type

@@

\begin{invariant}[\linketa \rhs]
  The \rhs of any \linketa in \linkStore has the shape $\lambda x.t_x$
  where $t_x$ is a \maybeeta term and $x$ is free in $t$.
  \label{inv:link-eta-right}
\end{invariant}

In the next three subsections we explain how we detect \maybeeta terms, how we
compile them and how the generated \linketa are activated during the \elpiIn{progress}.
Moreover, we provide justification for why \cref{inv:linklhs,inv:link-eta-right} remain true.

@@

Compiling a term $t$ forces us to determine if there exists a $t' \in \subterm{t}$,
such that $t' = \lambda x. t''_x$,
for any term $t''$, can be a $\eta$-expansion, i.e. under a
given substitution $\sigma$, we have $\sigma (\lambda x.t''_x) = t'$. This
\maybeeta detection is not a trivial operation as it may seems.
\\
\noindent
\begin{minipage}{.5\linewidth}%
  \begin{gather}
    \lambda x. f \appsep A_x\\
    \lambda x. f \appsep x \appsep A_x
  \end{gather}
\end{minipage}% This must go next to `\end{minipage}`
\begin{minipage}{.5\linewidth}
  \begin{gather}
    \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}\\
    \lambda x. f \appsep A_x \appsep x \label{eq:lamdisc}
  \end{gather}
\end{minipage}
